---
title: "U3L8 - Linear Classification"
output: html_document
date: "Type date here"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter
```{r}
#Delete all objects from Environment - Use with caution
remove(list = ls()) 

#Load Libraries
library(MASS) #For lda function (list before tidyverse so dplyr::select is not masked)
library(tidyverse)
library(gridExtra)
library(palmerpenguins)
library(caret) #For confusionMatrix() function
library(pROC) #For ROC Curves


#Read in Data
titanic <- read.csv("../../sample_data/L08_titanic3.csv")
```


## Example - Logistic Regression in R using Titanic Data
#### Part b - Predict Survived Using Age (This code produces an ERROR)
```{r, error = TRUE}
#Build Logistic Regression Model for Predicting Survived Using Age
#model1 <- glm(Survived ~ Age, family = binomial, data = titanic)
```


#### Part c - Predict Survived Using Age (Solution 1) 
```{r}
#CAUTION: Be sure to use the data from Canvas not the built in Titanic dataset
titanic <- titanic %>%
  mutate(SurvivedNum = ifelse(Survived == "Yes",1,0))
#Convert Survived to a 0/1 indicator
model1 <- glm(SurvivedNum ~ Age, family = binomial, data = titanic)
#Build Logistic Regression Model for Predicting SurvivedNum Using Age
summary(model1)
#Display a summary of the model

```

The estimated logit equation is given by:

$$log(\frac{\hat{p}_i}{1-\hat{p}_i})=-0.2092 - 0.0088x_{i,age}$$

The estimated probability form is given by:

$$\hat{p}_i=\frac{e^{-0.2092 - 0.0088x_{i,age}}}{1+e^{-0.2092 - 0.0088x_{i,age}}}$$

#### Part d - Predict Survived Using Age (Solution 2)
```{r}
#Build logistic regression using Survived as a factor
model1Factor <- glm(as.factor(Survived) ~ Age, family = binomial, data = titanic)
#Display a Summary of the Model
summary(model1Factor)

#What happens if Survived took values of "Did" and "Did Not"? Do answers change?
#Create New Variable that takes values of "Did" and "Did Not"
titanic <-
  titanic %>% mutate(Survived2 = ifelse(Survived == "Yes", "Did", "Did Not"))

#Build logistic regression model using alternative factor version of Survived
model1Factor2 <- glm(as.factor(Survived2) ~ Age, family = binomial, data = titanic)

#Display a Summary of the Model
summary(model1Factor2)
```

#### Part e - Check the factor levels
```{r}
#Check Levels
levels(as.factor(titanic$Survived))

levels(as.factor(titanic$Survived2))
```

#### Part h - Obtain predictions for a person who is 35 years old
```{r}
#Prediction for the LOGIT of a person who is 35
predict(model1, newdata = data.frame(Age=35))

# For probability
predict(model1, newdata = data.frame(Age=35), type="response")
```

#### Part i - Would you predict that a 35 year old survives or not?

No, the would not.

#### Part j - Obtain predictions for all people and the dataset; using a threshold of 0.38, predict whether each person survives or not
```{r}
predProb <- predict(model1, newdata = titanic, type = "response")

threshold <- 0.38

predSurv <- ifelse(predProb > threshold, "Yes", "No")
```


## Example - Assessing Predictive Performance of Titanic Model
#### Part b - Create the confusion matrix
```{r}
#Create the confusion matrix
table(predSurv, titanic$Survived)
```

#### Part d - Write code to find the accuracy 
```{r}
#Calculate the accuracy
sum(predSurv == titanic$Survived) / nrow(titanic)
mean(predSurv == titanic$Survived)
```



#### Part e - Write code to find the sensitivity
```{r}
#Find the number of true positives
tp <- sum(predSurv == "Yes" & titanic$Survived == "Yes")
tp
#Find the number of false negatives
fn <- sum(predSurv == "No" & titanic$Survived == "Yes")
fn
#Calculate the sensitivity
sen <- tp/(tp + fn)
sen
```

#### Part f - Use the confusionMatrix function from caret package (Incorrect Results)
```{r}
#Create confusion matrix using confusionMatrix function from caret library
confusionMatrix(data = as.factor(predSurv),
                reference = as.factor(titanic$Survived))
```

#### Part g - Use the confusionMatrix function from caret package (Corrected Results)
```{r}
#Create confusion matrix using confusionMatrix function from caret library
confusionMatrix(data = as.factor(predSurv),
                reference = as.factor(titanic$Survived),
                positive =  "Yes")
```

#### Part h - Create the ROC curve and find the area under the curve. Comment on the effectiveness of our model for making predictions.
```{r}
#Use the roc function from the pROC package to create the ROC curve
rocCurve <- roc(response = titanic$Survived,
                predictor = predProb,
                plot = TRUE,
                print.auc = TRUE,
                legacy.axes = TRUE)

#Extract the AUC value
as.numeric(rocCurve$auc)
```

The model is not doing a very good job.

#### Part i - Perform backward elimination, create the ROC curve, and find AUC
```{r}
#Build Intercept Only Model. NOTE: ~ 1 tells R that you only want an intercept
int_only_model <- glm(SurvivedNum ~ 1, 
                      family = binomial,
                      data = titanic)

#Build model with all potential regressors. 
#In code below, SurvivedNum ~ . tells R to use all columns in dataset to predict SurvivedNum
#SurvivedNum ~ . -Survived -Survived2 tells R to use all columns except Survived and Survived2 to predict SurvivedNum. If you get an error saying that object Survived2 cannot be found, delete it from the code below. 
full_model <- glm(SurvivedNum ~ . -Survived -Survived2, 
                  family = binomial,
                  data = titanic)

#Perform backward elimination
#Have R do it all
stats::step(object = full_model, 
            scope = list(lower = int_only_model, upper = full_model),
            data = titanic,
            direction = "backward")
```

```{r}
#Build model determined by backward elimination
bestModel <- glm(SurvivedNum ~ Pclass+Sex+Age+Siblings,
                  family = binomial,
                  data = titanic)

#Obtain Predictions
predProb <- predict(bestModel, newdata = titanic, type = "response")

#Create ROC curve
rocCurve2 = roc(response = titanic$Survived,
               predictor = predProb, 
               plot = TRUE, 
               print.auc = TRUE, 
               legacy.axes=TRUE)

#Extract the AUC
as.numeric(rocCurve2$auc)
```


## Example - Using LDA to Predict Species in R
#### Part a - Create copy of the dataset in your Environment and remove observations with missing values in species, bill length, and bill depth
```{r}
#Create a copy of the dataset in the Environment
data(penguins)

#Omit observations with missing values of species, bill length/depth
penguins <- na.omit(penguins %>% select(species, bill_length_mm, bill_depth_mm))
```

#### Part b - Check the normality assumption within each group
```{r}
#Distribution of Bill Length (all species together)
BLall <-
  ggplot(data = penguins, mapping = aes(x = bill_length_mm)) +
  geom_density() +
  labs(x = "Bill Length")

#Distribution of Bill Depth (all species together)
BDall <-
  ggplot(data = penguins, mapping = aes(x = bill_depth_mm)) +
  geom_density() +
  labs(x = "Bill Depth")

#Distribution of Bill Length (broken down by species)
BLeach <-
  ggplot(data = penguins, mapping = aes(x = bill_length_mm,
                                        fill = species)) +
  geom_density(alpha = 0.5) +
  labs(x = "Bill Length") + 
  theme(legend.position="bottom")

#Distribution of Bill Depth (broken down by species)
BDeach <-
  ggplot(data = penguins, mapping = aes(x = bill_depth_mm,
                                        fill = species)) +
  geom_density(alpha = 0.5) +
  labs(x = "Bill Depth") + 
  theme(legend.position="bottom")

grid.arrange(BDall, BLall, BLeach, BDeach, ncol = 2)
```

The first row of plots show that we're not following a normal distribution

The second row shows that the individual distributions for each species is normal.

There are no significant variables

#### Part c - Prepare the dataset for LDA
```{r}
#Scale the numeric features
xvars <- c("bill_length_mm", "bill_depth_mm")
penguins[ , xvars] <- scale(penguins[ , xvars], 
                            center = TRUE, 
                            scale = TRUE)

#Perform an 80/20 training/validation split using a seed of 123
set.seed(123)
trainInd <- sample(1:nrow(penguins), floor(0.8*nrow(penguins)))
set.seed(NULL)

Train <- penguins[trainInd, ]
Validation <- penguins[-trainInd,]
```

#### Part d - Perform LDA on the Train
```{r}
#Build LDA model
# comes from MASS
ldaModel <- lda(species ~ bill_length_mm + bill_depth_mm, data = Train)
ldaModel2 <- lda(Sex ~ Age + Fare, data = titanic)
#Display a summary (NOTE: we are NOT using summary())
ldaModel # This is how we get the summary
```


#### Part e - Generate and extract the posterior predicted probabilities and predicted classes for the Validation data.
```{r}
#Obtain Predictions for Validation Data
preds <- predict(ldaModel, newdata = Validation)

preds2 <- predict(ldaModel2, newdata = data.frame(Age = 35, Fare = 53.10))

round(preds2$posterior, digits = 3)
#preds is a list of 3. What elements are in preds?
names(preds)#Extract predicted probabilities
round(preds$posterior, digits = 3) # Generally do not display this
#Extract predicted classes (assign each observation to class with highest probability)
preds$class
```

#### Part f - Create the confusion matrix and calculate the accuracy
```{r}
#Create Confusion Matrix
table(preds$class, Validation$species)
#Calculate accuracy
mean(preds$class == Validation$species)
```