---
title: "U3L9 - Nonlinear Classification"
output: html_document
date: "Type date here"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter
```{r}
#Delete all objects from Environment - Use with caution
remove(list = ls()) 

#Load Libraries
library(tidyverse)
library(palmerpenguins)
library(FNN) #For k-nearest neighbors
library(rpart) #For classification and regression trees
library(rattle) #For creating nice visualizations of trees

#Read in Data
titanic <- read.csv("../../sample_data/L08_titanic3.csv")
```

## Example - k-Nearest Neighbors in R using Titanic Data
#### Part a - Prepare the data 
```{r}
#Check for NA's
apply(X = titanic, MARGIN = 2, FUN = function(x) sum(is.na(x)))   #margin = 2 means column while 1 means row
#Create Indicators
titanic <- 
  titanic %>% 
  mutate(Male = ifelse(Sex == "male", 1, 0))
  

#Create a vector to hold the names of input variables
xvars <- c("Age", "Male", "Fare", "Pclass")
#Scale the input variables
titanic[, xvars] <- scale(x = titanic[, xvars], center = TRUE, scale = TRUE)


#Perform Training/Validation Split
set.seed(123)
trainInd <- sample(1:nrow(titanic), floor(0.85 * nrow(titanic)))
set.seed(NULL)

Train <- titanic[trainInd, ]
Validation <- titanic[-trainInd, ]
```

#### Part b - Build the kNN model using 5 nearest neighbors 
```{r}
#Build kNN model with k = 5
knnRes <- knn(train = Train[, xvars, drop = FALSE],
              test = Validation[, xvars, drop =FALSE],
              cl = Train$Survived,
              k = 5) #drop = false prevents coercion down to a vector for 1 column dataframe
#View Results
#knnRes #Generally, do not include this in your code
```

#### Part c - Add the predictions to the validation set and name the new variable predSurvival.
```{r}
#Add Predictions to Validation Data (name them predSurvival)
Validation <-
  Validation %>%
  mutate(predSurvival = knnRes)
```


#### Part d - Create a confusion matrix
```{r}
#Create a confusion matrix (rows = predictions, columns = true status)
table(Validation$predSurvival, Validation$Survived)
```

#### Part e - Find the overall accuracy of the classifier
```{r}
#Calculate the accuracy
mean(Validation$Survived == Validation$predSurvival)
```


## Example - Determining the Optimal Number of Nearest Neighbors using a Loop
#### Part a - Use a for loop to run the kNN classifier for values of k from ranging 1 to 75
```{r}
#Initialize 
maxK <- 100
accVec <- rep(NA, maxK)

#Loop for using various values of k
for(i in 1:maxK){
  #Build kNN model
  knnRes <- knn(train = Train[, xvars, drop = FALSE],
              test = Validation[, xvars, drop =FALSE],
              cl = Train$Survived,
              k = i)
  #Add predictions to validation data
  Validation <- Validation %>%
      mutate(predSurvival = knnRes)
  
   accVec[i] <- mean(Validation$Survived == Validation$predSurvival)
  #Store the accuracy result
 
}

#View the accuracy values
which.max(accVec)
accVec[which.max(accVec)]
```

#### Part b - Create a plot showing the accuracy values as a function of k
```{r}
#Create a temporary data frame for plotting
tempDF <- data.frame(k = 1:maxK, Accuracy = accVec)

#Create a plot showing accuracy as a function of k
ggplot(tempDF, aes(x = k, y = Accuracy)) + geom_line() + geom_point(data = tempDF[which.max(accVec),], color = "red", shape = 1, size = 3)
```

#### Part c - Repeat the process using a random seed of 321. Which value of k would you choose?
```{r}
#Perform Training/Validation Split
set.seed(321)
trainInd <- sample(1:nrow(titanic), floor(0.85 * nrow(titanic)))
set.seed(NULL)

Train <- titanic[trainInd, ]
Validation <- titanic[-trainInd, ]

#Initialize 
maxK <- 75
accVec2 <- rep(NA, maxK)

#Loop for using various values of k
for(i in 1:maxK){
  #Build kNN model
  knnResTemp <- 
    knn(train = Train[ , xvars, drop = FALSE],
        test = Validation[ , xvars, drop = FALSE],
        cl = Train$Survived,
        k = i)
  
  #Add predictions to validation data
  Validation <-
    Validation %>%
    mutate(predSurvival = knnResTemp)
  
  #Store result
  accVec2[i] <- mean(Validation$Survived == Validation$predSurvival)
}

#View the accuracy values
accVec2

#Determine the optimal number of neighbors
which.max(accVec2)

#Determine the optimal accuracy
accVec2[which.max(accVec2)]

#Create a temporary data frame for plotting
tempDF2 <- data.frame(k = 1:maxK, Accuracy = accVec2)

#Create a plot showing accuracy as a function of k
ggplot(data = tempDF2,
       mapping = aes(x = k, y = Accuracy)) +
  geom_line() +
  labs(x = "Number of Nearest Neighbors",
       y = "Accuracy") +
  geom_point(data = tempDF2[which.max(accVec2), ],
             color = "red", shape = 1, size =3)
```

#### Part d - k-fold Cross Validation
```{r}
#Create fold values
num_folds <- 10
folds <- cut(x = 1:nrow(titanic), breaks = num_folds, labels = FALSE)

set.seed(123)
folds <- sample(folds) #randomly permute the fold values
set.seed(NULL)

#Initialize 
maxK <- 75
accMatrix <- matrix(NA, nrow = num_folds, ncol = maxK)

#Loop for using various values of k
for(i in 1:num_folds){
  #Perform Data Split
  valInd <- which(folds == i)
  Validation <- titanic[valInd, ]
  Train <- titanic[-valInd, ]
  
  for(j in 1:maxK){
    #Build kNN model
    knnResTemp <- 
    knn(train = Train[ , xvars, drop = FALSE],
        test = Validation[ , xvars, drop = FALSE],
        cl = Train$Survived,
        k = j)
    #Add predictions to validation data
    Validation <-
      Validation %>%
      mutate(predSurvival = knnResTemp)
    
    #Store result
    accMatrix[i,j] <- mean(Validation$Survived == Validation$predSurvival)
    }
}

#For each value of k, average accuracy values across the 10 folds
accMeans <- apply(accMatrix, 2, mean)

#Create a temporary data frame for plotting
tempDF <- data.frame(k = 1:maxK, Accuracy = accMeans)

#Create a plot showing accuracy as a function of k
ggplot(data = tempDF,
       mapping = aes(x = k, y = Accuracy)) +
  geom_line() +
  labs(x = "Number of Nearest Neighbors",
       y = "Accuracy") +
  geom_point(data = tempDF[which.max(accMeans), ],
             color = "red", shape = 1, size =3)
```


## Example - Classification Tree in R using Titanic Data
#### Part a - Perform an 85/15 training/validation split using a seed of 123.
```{r}
#Read Data into R in case original dataset was modified in earlier example
titanic <- read.csv("../../sample_data/L08_titanic3.csv")

#Perform Training/Validation Split
set.seed(123)
trainInd <- sample(1:nrow(titanic), floor(0.85 * nrow(titanic)))
set.seed(NULL)

Train <- titanic[trainInd, ]
Validation <- titanic[-trainInd, ]
```


#### Part b - Grow the classification tree. What are the first 2 splits?
```{r}
#Build the classification tree
classTree <- rpart(Survived  ~ ., data = Train, method = "class")
#If Survived was a 0/1 indicator, convert Survived to a factor to ensure Classification
#NOTE: method = "anova" was used for a regression tree

#Display a summary of the tree
summary(classTree) #Generally speaking, do not include this in your markdown

# The improvement is the based on the GINI index
```

First Split:

Second Split: 

#### Part c - Create a visualization of the tree. 
```{r}
#Visualize the tree (fancyRpartPlot is from the rattle package)
fancyRpartPlot(classTree, cex = .6)
```

#### Part e - Predict whether a 35 year-old female passenger who paid 50 for a class 2 ticket will survive 
```{r}
predict(classTree, newdata = data.frame(Age = 35, Sex = "female", Fare = 50, Pclass = 2, Siblings = 0, Parch = 0))
```

#### Part f - Obtain predicted probabilities for the validation set
```{r}
#Find Predicted Probabilities
predProb <- predict(classTree, newdata = Validation, type = "prob")
#Returns a matrix with 1 row per obs in Validation, and 1 column per response class

#View first six probability predictions
head(predProb, 6)
```

#### Part g - Obtain predicted classes for the validation set
```{r}
#Determine predicted classes
predSurv <- predict(classTree, newdata = Validation, type = "class")

#Show the first six predictions
head(predSurv)
```

NOTE: To obtain predictions for a regression tree, you can omit the `type = ` specification.

#### Part h - Obtain Confusion matrix and calculate accuracy
```{r}
#Create the confusion Matrix
table(predSurv, Validation$Survived)

#Calculate accuracy
mean(Validation$Survived == predSurv)
```

NOTE: For a binary response, we could use the predicted probabilities to generate an ROC curve and/or find AUC.

```{r}
rocCurve <- pROC::roc(response = Validation$Survived,
                      predictor = predProb[,2],
                      legacy.axes = TRUE,
                      print.auc = TRUE, plot = TRUE)
```


NOTE: For a regression tree, we could use the predicted responses to calculate MSE (or RMSE) for the validation set instead of the accuracy.


## Example - Determining Variable Importance
```{r}
summary(classTree)
```

The 3 most important features are

## Example - Node Size
Putting each observation into its own node would produce a Gini Index of 0, which suggests that each node is pure. Explain why this is not desirable. 


## Example -  Pruning the Tree
#### Part a - Using a complexity parameter (cp) value of 0.001, grow the tree. 
```{r}
#Grow the tree and perform 10-fold CV
set.seed(123)
fullTree <- rpart(Survived ~ . , data = Train, method = "class", cp = 0.001, xval = 10)

#Display the cptable
fullTree$cptable
```

#### Part b - Plot the tree
```{r}
fancyRpartPlot(fullTree, cex = 0.7)
```

#### Part c - Visualize the results of the cross validation procedure. 
```{r}
plotcp(fullTree)
```

#### Part d - What is the optimal value of the complexity parameter based on minimizing the cross validation error? 
```{r}
#Determine the row of cptable containing smallest CV error
cpRow <- which.min(fullTree$cptable[ , "xerror"])

#Determine optimal value of cp 
cpChoiceMin <- fullTree$cptable[ , "CP"][cpRow]

#Display the choice
cpChoiceMin
```

#### Part e - What is the optimal value of the complexity parameter based on the 1se rule?
```{r}
#Determine the row of cptable containing smallest CV error
cpRow <- which.min(fullTree$cptable[ , "xerror"])

#Calculate min(xerror) + xstd
target <- fullTree$cptable[ , "xerror"][cpRow] +fullTree$cptable[ , "xstd"][cpRow]

#Determine which xerror values are less than target and select first such value
cpRow1se <- which(fullTree$cptable[ , "xerror"] < target)[1]

cpChoice1se <- fullTree$cptable[ , "CP"][cpRow1se]

#Display the choice
cpChoice1se
```

#### Part f - Prune the tree
```{r}
#Prune the true
prunedTree <- prune(fullTree, cp=cpChoice1se)

#View the pruned tree
fancyRpartPlot(prunedTree, cex=0.65)
```

#### Part g - Using the pruned tree, find the validation set accuracy
```{r}
#Obtain the predicted classes
prunedPred <- predict(prunedTree, newdata = Validation, type = "class")
  
#Calculate the accuracy
mean(Validation$Survived == prunedPred)
```