---
title: "U1L1 - Data Viz and Data Wrangling"
output: html_document
date: "August 21, 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter
```{r}
#Delete all objects from Environment - Use with caution
remove(list = ls()) 

#Load Libraries
library(tidyverse)
library(palmerpenguins)

#Load Datasets

```

## Example - Common visualizations

The following illustrates some common types of plots and the corresponding code. The examples require the `dplyr` and `ggplot2` libraries, both of which are loaded as part of the `tidyverse`. The examples also `penguins` data set from the `palmerpenguins` library.

```{r}
ggplot(data = penguins, mapping = aes(x = species)) +
  geom_bar(fill = "white", color = "black") +
  labs(x = "Species",
       y = "Number of Penguins",
       title = "Penguin Species in Palmer Penguins Dataset")

#NOTE: Inside aes(), no y axis is specified. This is because the y axis is not a variable contained in the dataset. In a bar plot, the y axis is the count which is determined by the geom_bar procedure
```

```{r}
ggplot(data = starwars, mapping = aes(x = height)) + 
  geom_histogram(binwidth = 20, color = "black", fill = "white") +
  labs(x = "Height (in Centimeters)",
       y = "Count",
       title = "Character Heights in Star Wars")
#NOTE: Inside aes(), no y axis is specified. This is because the y axis is not a variable contained in the dataset. In a histogram, the y axis is the count which is determined by the geom_histogram procedure
```

```{r}
ggplot(data = starwars, mapping = aes(x = height)) + 
  geom_density() +
  labs(x = "Height (in Centimeters)",
       y = "Density",
       title = "Character Heights in Star Wars")
```

```{r}
ggplot(data = starwars, mapping = aes(x = height, y = mass)) +
  geom_point(shape = 5) +
  labs(title = "Weight vs. Height for Characters in Star Wars",
       x = "Height (in Centimeters)", y = "Weight (in Kilograms)")
```

```{r}
ggplot(data = penguins, mapping = aes(x = sex, y = body_mass_g)) +
  geom_boxplot() +
  labs(title = "Body Mass vs. Sex for Palmer Penguins",
       x = "Sex of Penguin", 
       y = "Body Mass (in Grams)")
```

## Example - Loading and previewing a dataset from a package
```{r}
data(penguins)
glimpse(penguins)
```

## Example - Layers of ggplot demonstration
```{r}

```










## Example - Using aesthetics
```{r}

```


## Example - Faceting using two categorical variables
```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) + 
  geom_point(shape = 5) +
  facet_grid(species ~ sex) +
  labs(x = "Bill Length (in mm)",
       y = "Bill Depth (in mm)")
```

## Example - Faceting using one categorical variable
```{r}
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) + 
  geom_point(shape = 5) +
  facet_grid(. ~ species) +
  labs(x = "Bill Length (in mm)",
       y = "Bill Depth (in mm)")

```

## Example - Supporting data visualization with numerical summaries

#### Part a 

#### Part b - Incorrect Code

```{r}
penguins %>%
  group_by(species) %>%
  summarize(N = n(),
            MeanMass = mean(body_mass_g),
            MedianMass = median(body_mass_g),
            StdDevMass = sd(body_mass_g))
```

#### Part b - Corrected
```{r}
penguins %>%
  group_by(species) %>%
  summarize(N = n(),
            MeanMass = mean(body_mass_g),
            MedianMass = median(body_mass_g),
            StdDevMass = sd(body_mass_g))
```

## Part c


## Example - Alternative methods for finding the summary statistics
```{r}
#Alternative Method

#Create dataset for each species and calculate values; repeat for each species
df1 <- filter(penguins, species == "Adelie")
nrow(df1)
mean(df1$body_mass_g, na.rm = TRUE)
median(df1$body_mass_g, na.rm = TRUE)
sd(df1$body_mass_g, na.rm = TRUE)

df2 <- filter(penguins, species == "Chinstrap")
nrow(df2)
mean(df2$body_mass_g, na.rm = TRUE)
median(df2$body_mass_g, na.rm = TRUE)
sd(df2$body_mass_g, na.rm = TRUE)    

df3 <- filter(penguins, species == "Gentoo")
nrow(df3)
mean(df3$body_mass_g, na.rm = TRUE)
median(df3$body_mass_g, na.rm = TRUE)
sd(df3$body_mass_g, na.rm = TRUE)
```


## Example - dplyr practice in R

#### Part a
```{r}

```


#### Part b
```{r}

```
















