---
title: "STAT508_DAA6"
author: "Ada Lazuli"
output: html_document
date: "Thursday, Feb 26, 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter

```{r, warning=FALSE, message=FALSE}
remove(list = ls()) 
library(AmesHousing)
library(tidyverse)
library(ISLR2)
library(pls)
library(reactable) 
library(glmnet)
```

## Code for Creating the Processed Dataset
```{r}
ames <- AmesHousing::make_ames()
dim(ames)
```

## Problem 1 - Prepare the Dataset for Analysis
#### Part a. - Subset to only include the numeric variables and remove observations with missing values

```{r}
ames2 <- ames %>% select(where(is.numeric))
ames2 <- na.omit(ames2)
dim(ames2)
```

#### Part b. - Perform a training/validation split

```{r}
set.seed(123)
trainInd <- sample(1:nrow(ames2), floor(0.8 * nrow(ames2)), replace = FALSE)
train <- ames2[trainInd,]
validation <- ames2[-trainInd,]
```


## Problem 2 - Perform Principal Component Regression (PCR)
#### Part a. - Fit a PCR model and display a summary of the results. Use 10-fold CV.

```{r}
set.seed(1)
pcrFit <- pcr(Sale_Price ~ ., data = train, center = TRUE, scale = TRUE, validation = "CV", segments = 10) 
```


#### Part b. - Create a plot showing the cross validation RMSE values. How many components? 

```{r}
tempRMSE <- RMSEP(pcrFit)
ggplotDF <- data.frame(Components = 0:(ncol(train)-1), 
                       RMSE = tempRMSE$val[1, , ]) 
optMpcr <- which.min(tempRMSE$val[1, , ]) 

ggplot(data = ggplotDF, mapping = aes(x = Components, y = RMSE)) +
  geom_line() +
  geom_point() +
  geom_point(data = ggplotDF[optMpcr,], color = "red", shape = 1, size = 4) +
  scale_x_continuous(breaks = seq(0, ncol(train)-1, 2)) +
  labs(x = "Number of Components",
       y = "RMSE (from 10-fold CV)")
```

Based on the training cross fold valdation RSME, the optimal number of components is 31.


#### Part c. - Based on the optimal number of principal components chosen in the Part b., what percent of the variance in Sale_Price is explained by the PCR model using the training data?

```{r}
summary(pcrFit)
```

Based on the summary, 80.15% of the variation in `sale_price` is explained by the first 31 components.

#### Part d. - Compute the validation set mean squared error (MSE) and RMSE based on PCR with the optimal number of components.

```{r}
yhatPCR <- predict(pcrFit, newdata = validation, ncomp = 31)
MSEpcr <- mean((validation$Sale_Price - yhatPCR)^2)
RMSEpcr <- sqrt(MSEpcr)
RMSEpcr
```

The RMSE score against the validation set is **`r format(round(RMSEpcr, digits = 2), scientific = FALSE)`** with the PCR model

## Problem 3 - Perform Partial Least Squares Regression
#### Part a. - Fit a PLS regression model and display a summary of the results. Use 10-fold CV.

```{r}
set.seed(1)
plsFit <- plsr(Sale_Price ~ ., data = train,
               center = TRUE, scale = TRUE,
               validation = "CV", segments = 10)
```

#### Part b. - Create a plot showing the cross validation RMSE values. How many components?  

```{r}
tempRMSE <- RMSEP(plsFit)

ggplotDF <- data.frame(Components = 0:(ncol(train)-1), 
                       RMSE = tempRMSE$val[1, , ])

optMpls <- which.min(tempRMSE$val[1, , ]) 
ggplot(data = ggplotDF, mapping = aes(x = Components, y = RMSE)) +
  geom_line() +
  geom_point() +
  geom_point(data = ggplotDF[optMpls,], color = "red", shape = 1, size = 4 ) +
  scale_x_continuous(breaks = seq(0, ncol(train)-1, 2)) +
  labs(x = "Number of Components",
       y = "RMSE (from 10-fold CV)")
```

Based on the RMSE score over 10 cross fold validations on the training set, it appears that using 2 components is optimal for PLS.

#### Part c. - Based on the optimal number of principal components chosen in the Part b., what percent of the variance in Sale_Price is explained by the PLS model using the training data?

```{r}
summary(plsFit)
```

Two components in the PLS model explain 77.94% of the variance in `Sale_Price`

#### Part d. - Compute the validation set mean squared error (MSE) and RMSE based on PLS with the optimal number of components.

```{r}
yhatPLS <- predict(plsFit, newdata = validation, ncomp = 2)
MSEpls <- mean((validation$Sale_Price - yhatPLS)^2)
RMSEpls <- sqrt(MSEpls)
RMSEpls
```

The RMSE score against the validation set is  **`r format(round(RMSEpls, digits = 2), scientific = FALSE)`**  with the PLS model

## Problem 4 - Comparing Methods
#### Part a. - Ordinary Least Square Linear Regression

```{r}
lrModel <- lm(Sale_Price ~ ., data = train)
yhatLR <- predict(lrModel, newdata = validation)
MSElr <- mean((validation$Sale_Price - yhatLR)^2)
RMSElr <- sqrt(MSElr)
RMSElr
```

#### Part b. - LASSO Regression

```{r}
# Create the matrices for LASSO and Ridge
Xmat <- model.matrix(Sale_Price ~ ., data = ames2)[, -1]
yvec <- ames2$Sale_Price

set.seed(123)
trainInd <- sample(1:nrow(ames2), floor(0.8*nrow(ames2)), replace = FALSE)
set.seed(NULL)
XmatTrain <- Xmat[trainInd,]
XmatVal <- Xmat[-trainInd,]
yvecTrain <- yvec[trainInd]
yvecVal <- yvec[-trainInd]

set.seed(1)
lassoCV <- cv.glmnet(x = XmatTrain, y = yvecTrain,
                    family = "gaussian",
                    alpha = 1, 
                    lambda = NULL,
                    standardize = TRUE,
                    nfolds = 10)
lassoYhat <- predict(lassoCV, s = lassoCV$lambda.min, newx = XmatVal) 
lassoMSE <-  mean((yvecVal - lassoYhat)^2)
lassoRMSE <- sqrt(lassoMSE)
lassoRMSE
```

#### Part c. - Ridge Regression

```{r}
set.seed(1)
ridgeCV <- cv.glmnet(x = XmatTrain, y = yvecTrain,
                    family = "gaussian",
                    alpha = 0, 
                    lambda = NULL,
                    standardize = TRUE,
                    nfolds = 10)
ridgeYhat <- predict(ridgeCV, s = ridgeCV$lambda.min, newx = XmatVal) 
ridgeMSE <-  mean((yvecVal - ridgeYhat)^2)
ridgeRMSE <- sqrt(ridgeMSE)
ridgeRMSE
```

#### Part d. - Which method worked the best? 

```{r}
performance_df <- data.frame(Method = c("Linear Regression (All Features)", "PCR (31 Components)", "PLS (2 Components)", "LASSO", "Ridge"),
                             RMSEValidation = c(RMSElr, RMSEpcr, RMSEpls, lassoRMSE, ridgeRMSE))

performance_df <- performance_df %>% arrange(RMSEValidation) %>% mutate(RMSEValidation = round(RMSEValidation, digits = 2))

reactable(performance_df)
```

Based on the RMSE score against the validation set, it seems that Ridge regression performs the best against with held data it was not exposed to during training and 
it would follow that the Ridge model would perform the best against new observations in production.


