---
title: "U2L6 - Dimension Reduction in Regression"
output: html_document
date: "Type date here"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter
```{r}
#Delete all objects from Environment - Use with caution
remove(list = ls()) 

#Load Libraries
library(tidyverse)
library(ISLR2)
library(pls) #used for building PCR and PLS models
```

## Motivating Example
#### Simulate a dataset
```{r}
# Set random seed for reproducibility
set.seed(123)

# Number of observations and predictor variables
n <- 50  # Number of observations
p <- 75  # Number of predictor variables

# Create random predictor variables
X <- matrix(rnorm(n * p), nrow = n)

# Create random coefficients for the predictor variables (no intercept)
beta <- rnorm(p)

# Create the response variable as a linear combination of predictor variables (including an intercept of 3)
Y <- 3 + X %*% beta + rnorm(n)

# Create a dataframe for the dataset
df <- data.frame(Response = Y, as.data.frame(X))
```


#### Part a. - Linear regression on Simulated Dataset
```{r}
model1 <- lm(Response ~ ., data = df)
summary(model1)
```


## Example - Applying PCR in R
#### Part a. - Prepare the data
```{r}
#Create a copy of the Hitters dataset in your Environment
data(Hitters)

#Omit any player that has a missing value in at least one variable
Hitters <- na.omit(Hitters)

#Perform an 80/20 training/validation split using a seed of 123
set.seed(123)
trainInd <- sample(1:nrow(Hitters), floor(0.8 * nrow(Hitters)), replace = FALSE)
set.seed(NULL)

#Create training and validation datasets
train <- Hitters[trainInd,]
validation <- Hitters[-trainInd,]

```

#### Part b. - Understanding the algorithm
```{r}
#Perform PCA on the Input centered and scaled input data
#Prepare the data
Xmat <- model.matrix(Salary ~ ., data = train)[, -1] # This will create the indicator variables
#Perform PCA
praRes <- prcomp(x = Xmat, center = TRUE, scale. = TRUE)

#Check Proportion of Total Variance Explained
# protion of variance explained
PVE <- praRes$sdev^2 / sum(praRes$sdev^2)
PVE
cPVE <- cumsum(PVE)
cPVE
#Use the 1st 2 PCs as the inputs into a linear regression model for Salary
#Create a data frame (tempdf) containing salary and the scores for first 2 PC's
tempdf <- data.frame(Salary = train$Salary, PC1 = praRes$x[,1], PC2 = praRes$x[,2])
#Build a linear regression model
linReg2PC <- lm(Salary ~ PC1 + PC2, data = tempdf)
summary(linReg2PC)
```

- What proportion of the total variance is explained using the first 2 PCs? `r cPVE[2]`

- What proportion of the total variation in Salary is explained by the linear regression model using the first 2 PCs? Multiple R-squared: 0.4629


#### Part c. - Fit the a PCR model using the pcr() function
```{r}
#Perform PCR
set.seed(1)
pcrFit <- pcr(Salary ~ ., data = train, center = TRUE, scale = TRUE, validation = "CV", segments = 10) # use "LOOV" for leave one out
set.seed(NULL)
summary(pcrFit)
```


#### Part d. - Create a plot showing the cross validation error. 
```{r}
#Extract RMSE values for the cross validation procedure
tempRMSE <- RMSEP(pcrFit)

#Create object (ggplotDF) for visualizing
ggplotDF <- data.frame(Components = 0:19, 
                       RMSE = tempRMSE$val[1, , ]) #Use [2, , ] for adjusted CV values

#Determine the optimal (lowest) CV RMSE
optMpcr <- which.min(tempRMSE$val[1, , ]) 

#Create Plot
ggplot(data = ggplotDF, mapping = aes(x = Components, y = RMSE)) +
  geom_line() +
  geom_point() +
  geom_point(data = ggplotDF[optMpcr,], color = "red", shape = 1, size = 4) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  labs(x = "Number of Components",
       y = "RMSE (from 10-fold CV)")
```


#### Part e. - Compute the validation set RMSE
```{r}
#Obtain predictions based on the optimal number of components
yhatPCR <- predict(pcrFit, newdata = validation, ncomp = 16)
#Calculate MSE and RMSE
MSEpcr <- mean((validation$Salary - yhatPCR)^2)
RMSEpcr <- sqrt(MSEpcr)
RMSEpcr
```

## Example - Applying PLS in R
#### Part a. - Prepare the data
```{r}
#####There is really no need to run this code as long as you have trainInd, train, and validation in your Environment from the PCR example#######

#Create a copy of the Hitters dataset in your Environment
data(Hitters)

#Omit any player that has a missing value in at least one variable
Hitters <- na.omit(Hitters)

#Perform an 80/20 training/validation split using a seed of 123
set.seed(123)
trainInd <- sample(1:nrow(Hitters), floor(0.8*nrow(Hitters)), replace = FALSE)
set.seed(NULL)

train <- Hitters[trainInd, ]
validation <- Hitters[-trainInd, ]
```


#### Part b. - Fit the a PLS model using the plsr() function
```{r}
#Perform PLS
set.seed(1)
plsFit <- plsr(Salary ~ ., data = train,
               center = TRUE, scale = TRUE,
               validation = "CV", segments = 10) # segmentation = folds
summary(plsFit)
```

#### Part c. - Create a plot showing the cross validation error. 
```{r}
#Extract RMSE values for the cross validation procedure
tempRMSE <- RMSEP(plsFit)

#Create object (ggplotDF) for visualizing
ggplotDF <- data.frame(Components = 0:19, 
                       RMSE = tempRMSE$val[1, , ]) #Use [2, , ] for adjusted CV values

#Determine the optimal (lowest) CV RMSE
optMpls <- which.min(tempRMSE$val[1, , ]) 

#Create Plot
ggplot(data = ggplotDF, mapping = aes(x = Components, y = RMSE)) +
  geom_line() +
  geom_point() +
  geom_point(data = ggplotDF[optMpls,], color = "red", shape = 1, size =4 ) +
  scale_x_continuous(breaks = seq(0, 20, 2)) +
  labs(x = "Number of Components",
       y = "RMSE (from 10-fold CV)")
```


#### Part d. - Compute the validation set RMSE
```{r}
#Obtain predictions based on the optimal number of components
yhatPLS <- predict(plsFit, newdata = validation, ncomp = 8)
#Calculate MSE and RMSE
MSEpls <- mean((validation$Salary - yhatPLS)^2)
RMSEpls <- sqrt(MSEpls)
RMSEpls
```

#### Part e. - Compare the methods
```{r}
#Fit a regular linear regression model using least-squares
linRegModel <- lm(Salary ~ ., data = train)
#Obtain Predictions
yhatLR <- predict(linRegModel, newdata = validation)
#Calculate MSE and RMSE
MSElr <- mean((validation$Salary - yhatLR)^2)
RMSElr <- sqrt(MSElr)
#Build data frame for easy comparison
df <- data.frame(Method = c("Linear Regression (All Features)", "PCR (16 Components)", "PLS (8 Components)"), RMSEValidation = c(RMSElr, RMSEpcr, RMSEpls))
df
```

