---
title: "U2L5 - Feature Selection and Regulariation"
output: html_document
date: "Type date here"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter
```{r}
#Delete all objects from Environment - Use with caution
remove(list = ls()) 

#Load Libraries
library(tidyverse)
library(gridExtra)
library(ISLR2)
library(leaps) #Not used before (may need to install)
library(glmnet) #Not used before (may need to install); used for LASSO/Ridge

```

## Example - Illustrating the Bias-Variance Tradeoff
#### User defined functions for this problem
```{r}
f = function(x) {
  #This is the true regression function
  x ^ 2
}

gen_sim_data = function(f, sample_size = 100) {
  #This function randomly generates a dataset of sample_size observations using y = f(x) + epsilon
  x = runif(n = sample_size, min = 0, max = 1) #randomly generate x's
  eps = rnorm(n = sample_size, mean = 0, sd = sqrt(0.5625)) #randomly generate noise
  y = f(x) + eps #generate y's using the regression form
  tibble(x, y)
}

get_mse <- function(truth, estimate) {
  #This function calculates the MSE 
  mean((estimate - truth) ^ 2)
}

get_bias <- function(estimate, truth) {
  #This function calculates the bias
  mean(estimate) - truth
}

get_var <- function(estimate) {
  #This function calculates the variance 
  mean((estimate - mean(estimate)) ^ 2)
}
```

#### Randomly generate a dataset; plot dataset and the true regression form
```{r}
#Randomly Generate a Dataset
set.seed(1)
sim_data <- gen_sim_data(f)

#Plot the dataset and the true regression function
SimulatedData <- 
  ggplot(data = sim_data, mapping = aes(x = x, y = y)) +
  geom_point(alpha = 0.35, size = 2) +
  geom_line(data = data.frame(x = seq(0, 1, 0.001), 
                              y = f(seq(0, 1, 0.001))),
            mapping = aes(x = x, y = y),
            linewidth = 1)

SimulatedData
```

#### Plot Each estimated line
```{r}
#Fit the models
fit_0 = lm(y ~ 1,                   data = sim_data)
fit_1 = lm(y ~ poly(x, degree = 1), data = sim_data)
fit_2 = lm(y ~ poly(x, degree = 2), data = sim_data)
fit_9 = lm(y ~ poly(x, degree = 9), data = sim_data)

#Create plots of the estimated models
#Plot of Degree 0 model
degree0 <-
  SimulatedData + 
  geom_line(data = data.frame(x = seq(0,1,0.001), y = predict(fit_0, data.frame(x=seq(0,1,0.001)))), color = "orange", linetype = "dashed",
            linewidth = 1) +
  labs(title = "Degree 0")

degree0

degree1 <-
  SimulatedData + 
  geom_line(data = data.frame(x = seq(0,1,0.001), y = predict(fit_1, data.frame(x=seq(0,1,0.001)))), color = "red", linetype = "dashed",
            linewidth = 1) +
  labs(title = "Degree 1")

degree1

degree2 <-
  SimulatedData + 
  geom_line(data = data.frame(x = seq(0,1,0.001), y = predict(fit_2, data.frame(x=seq(0,1,0.001)))), color = "blue", linetype = "dashed",
            linewidth = 1) +
  labs(title = "Degree 2")

degree2

degree9 <-
  SimulatedData + 
  geom_line(data = data.frame(x = seq(0,1,0.001), y = predict(fit_9, data.frame(x=seq(0,1,0.001)))), color = "purple", linetype = "dashed",
            linewidth = 1) +
  labs(title = "Degree 9")

degree9
```

#### Create one plot showing all four models
```{r}
gridExtra::grid.arrange(degree0, degree1, degree2, degree9, ncol = 2)
```

#### Illustrating Variance 
```{r}
#Illustrate Variance for Highly Flexible (degree = 9) Model
set.seed(123)
plot_list <- list()
for(i in 1:4){
  sim_data <- gen_sim_data(f)
  
  fit_temp = lm(y ~ poly(x, degree = 9), data = sim_data)
  
  plot_list[[i]] <-
  ggplot(sim_data, mapping = aes(x=x, y = y)) +
    geom_point(alpha = 0.35) + 
    geom_line(data = data.frame(x = seq(0,1,0.001), y = predict(fit_temp, data.frame(x=seq(0,1,0.001)))), color = "purple", linewidth = 1) +
    geom_segment(x = 0.9, y = 0, xend = 0.9,
                 yend = predict(fit_temp, data.frame(x = 0.9)),
                 color = "red", linewidth = 1) +
  labs(title = paste("Dataset ", i, sep = ""))
}

#Display results
gridExtra::grid.arrange(plot_list[[1]], plot_list[[2]], plot_list[[3]], plot_list[[4]],ncol = 2)

#Illustrate Variance for Highly Inflexible (degree = 0) Model
set.seed(123)
plot_list <- list()
for(i in 1:4){
  sim_data <- gen_sim_data(f)
  
  fit_temp = lm(y ~ 1, data = sim_data)
  
  plot_list[[i]] <-
  ggplot(sim_data, mapping = aes(x=x, y = y)) +
    geom_point(alpha = 0.35) + 
    geom_line(data = data.frame(x = seq(0,1,0.001), y = predict(fit_temp, data.frame(x=seq(0,1,0.001)))), color = "orange", linewidth = 1) +
    geom_segment(x = 0.9, y = 0, xend = 0.9,
                 yend = predict(fit_temp, data.frame(x = 0.9)),
                 color = "red", linewidth = 1) +
  labs(title = paste("Dataset ", i, sep = ""))
}

#Display the results
gridExtra::grid.arrange(plot_list[[1]], plot_list[[2]], plot_list[[3]], plot_list[[4]],ncol = 2)
```

#### Simulation Study
```{r}
#Initialize
set.seed(1)
n_sims <- 300
n_models <- 4
x <-  data.frame(x = 0.90) # fixed point at which we make predictions
predictions <- matrix(0, nrow = n_sims, ncol = n_models)

for (sim in 1:n_sims) {
  
  # simulate new, random, training data
  # this is the only random portion of the bias, var, and mse calculations
  # this allows us to calculate the expectation over D
  sim_data <- gen_sim_data(f)
  
  # fit models
  fit_0 <- lm(y ~ 1,                   data = sim_data)
  fit_1 <- lm(y ~ poly(x, degree = 1), data = sim_data)
  fit_2 <- lm(y ~ poly(x, degree = 2), data = sim_data)
  fit_9 <- lm(y ~ poly(x, degree = 9), data = sim_data)
  
  # get predictions
  predictions[sim, 1] <- predict(fit_0, x)
  predictions[sim, 2] <- predict(fit_1, x)
  predictions[sim, 3] <- predict(fit_2, x)
  predictions[sim, 4] <- predict(fit_9, x)
}

#Convert from wide form to long form for plotting
predictions_long <- 
  as.data.frame(predictions) %>%
  pivot_longer(cols = c(V1, V2, V3, V4), names_to = "degree", names_prefix = "V") %>%
  mutate(degree = case_when(degree == "1" ~ "0",
                            degree == "2" ~ "1",
                            degree == "3" ~ "2",
                            degree == "4" ~ "9"))

#Create plot showing predicted values for x0 = 0.9
set.seed(123) #This seed is for random jitter
ggplot(data = predictions_long, mapping = aes(x = degree, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(color = degree), alpha = 0.5, shape = 1, position = "jitter") +
  geom_hline(yintercept = f(0.9), color = "black", linetype = "dashed") +
  labs(x = "Degree of Polynomial", 
       y = "Prediction") + 
  theme(legend.position="none")



#Calculate Summaries of simulation study
bias <- apply(predictions, 2, get_bias, truth = f(x = 0.90))
variance <- apply(predictions, 2, get_var)
mse <- apply(predictions, 2, get_mse, truth = f(x = 0.90))

#Display the results
sim_results <- data.frame(Degree = c(0,1,2,9),
                          MSE = mse,
                          Bias = bias,
                          BiasSquared = bias^2,
                          Variance = variance)

round(sim_results, digits = 4)
#cbind(mse,bias^2,variance)

```


## Example - Best Subsets Using Indirect Estimates of Test Error
#### Part a
```{r}
#Create a copy of the dataset in the Environment
data("Credit")

#Perform Best Subsets
regfitBS <- regsubsets(Balance ~ ., data = Credit, nvmax = 11)

#Display the summary
summary(regfitBS)
```

#### Part b - Which variables are included in the best 3 variable model?

The variables included in the best 3 variable model are:

Income, Rating, StudentYes

#### Part c - Store the summary of the best subset selection process
```{r}
#Store the summary
regfitBSsummary <- summary(regfitBS)
```

#### Part d - View, pick the best model based on BIC, plot BIC, and find variables in the model
```{r}
#Part i - View the BIC values
 #similar options for rsq, rss, adjr2, and cp
regfitBSsummary$bic
#Part ii - Determine the size of the model associated with the best (smallest) BIC value
optSizeBIC <- which.min(regfitBSsummary$bic)
optSizeBIC
#Part iii - Plot the BIC values
#First create a data frame
BICdf <- data.frame(Model = 1:nrow(regfitBSsummary$outmat), BIC = regfitBSsummary$bic)
  
#Create a plot
ggplot(data = BICdf, mapping = aes(x = Model,y = BIC)) + geom_line() + geom_point() +
  scale_x_continuous(breaks = c(1:nrow(regfitBSsummary$outmat))) + 
  geom_point(data = BICdf[optSizeBIC, ], color = "red", size = 3) +
  labs(x = "Num Features", y = "BIC")
  
#Part iv - Which variables remain?
coef(regfitBS, id = optSizeBIC)
```

#### Part e - View, pick the best model based on Adjusted $R^2$, plot Adjusted $R^2$, and find variables in the model
```{r}
#Part i - View the Adjusted R^2 values
regfitBSsummary$adjr2 #similar options for rsq, rss, adjr2, and cp

#Part ii - Determine the size of the model associated with the best (largest) Adjusted R^2 value
optSizeAdjR2 <- which.max(regfitBSsummary$adjr2)
optSizeAdjR2

#Part iii - Plot the Adjusted R^2 values
#First create a data frame
AdjR2df <- data.frame(Model = 1:nrow(regfitBSsummary$outmat),
                    AdjR2 = regfitBSsummary$adjr2)


#Create a plot
ggplot(data = AdjR2df, mapping = aes(x = Model, y = AdjR2)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = c(1:11)) + 
  labs(x = "Number of Features in the Model",
       y = expression("Adjusted " * R^2 * " Value")) + #expression() is for getting exponent of 2 into title
  geom_point(data = AdjR2df[optSizeAdjR2,] , color = "red", size = 3)

#Part iv - Which variables remain?
coef(regfitBS, id = optSizeAdjR2)
```

## Example - Best Subsets Using Direct Estimates of Test Error (Validation Set Approach)
#### Part a - 80/20 training/validation split using a seed of 123
```{r}
#Random select 80% of the row numbers to identify the training set
set.seed(123)
trainInd <- sample(1:nrow(Credit), floor(0.8*nrow(Credit)), replace = FALSE)

#Create training and validation sets
train <- Credit[trainInd, ]
validation <- Credit[ -trainInd,]
```

#### Part b - Use the best subsets approach to identify the best model of each size
```{r}
#Apply the best subsets appraoch the training data
regfitBSval <- regsubsets(Balance ~ ., nvmax = 11, data = train)

#Display the summary
summary(regfitBSval)
```

#### Part c - Create and display the data matrix for validation set
```{r}
#Create data matrix
validationMat <- model.matrix(Balance ~ ., data = validation)

#Display first 6 rows
head(validationMat, 6)
```

#### Part d - Use a loop to obtain predictions and MSE for validation set for optimal model of each size
```{r}
#Initialize 
numModels <- ncol(validationMat) - 1 
valMSE <- rep(NA, numModels) #Repeats NA numModels times (11 in this problem)

#Loop to cycle through the models of different sizes
for (i in 1:numModels){
  coefi <- coef(regfitBSval, id = i) #Extract coefficients for model with i variables
  yhat <- validationMat[, names(coefi)] %*% coefi #yhat = X %*% betahat
  valMSE[i] <- mean((validation$Balance - yhat)^2) #Calculate and store val. set mse
}

#Display validation set MSE values
valMSE

#Determine which size produced the optimal
optSizeVal <- which.min(valMSE)
optSizeVal

#Create a plot of Validation Set MSE's
#First create a data frame
valMSEdf <- data.frame(Model = 1:numModels,
                       MSE = valMSE)

#Create a plot of Validation Set MSE's
ggplot(data = valMSEdf, mapping = aes(x = Model, y = MSE)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = c(1:11)) + 
  labs(x = "Number of Features in the Model",
       y = "Validation Set MSE") +
  geom_point(data = valMSEdf[optSizeVal,] , color = "red", size = 3)

#Display the coefficients of the optimal model (coefficients were estimated using train)
coef(regfitBS, id = optSizeVal)
```


## Example - Applying Shrinkage Methods in R
#### Part a - Preparing the data for glmnet
```{r}
#Create a copy of Credit in the Environment
data("Credit")

#Create input matrix and response vector
Xmat <- model.matrix(Balance ~ ., data = Credit)[, -1]
yvec <- Credit$Balance


#Perform training/validation split
set.seed(1)
trainInd <- sample(1:nrow(Credit), floor(0.8*nrow(Credit)), replace = FALSE)

set.seed(NULL)
#Split Xmat and yvec separately
XmatTrain <- Xmat[trainInd,]
XmatVal <- Xmat[-trainInd,]
yvecTrain <- yvec[trainInd]
yvecVal <- yvec[-trainInd]

```

#### Part b - Build the LASSO model and plot the coefficient paths
```{r}
#Fit the LASSO model (let R choice lambda sequence)
lassoModel <- glmnet(x = XmatTrain, y = yvecTrain,
                    family = "gaussian",
                    alpha = 1,
                    lambda = NULL,
                    standardize = TRUE)

#Create a plot of the coefficient paths
plot(lassoModel, xvar = "lambda", label = TRUE)
```

#### Part c - Build the Ridge model and plot the coefficient paths
```{r}
#Fit the Ridge model (let R choice lambda sequence)
ridgeModel <- glmnet(x = XmatTrain, y = yvecTrain,
                     family = "gaussian",
                     alpha = 0, 
                     lambda = NULL,
                     standardize = TRUE)

#Create a plot of the coefficient paths
plot(ridgeModel, xvar = "lambda", label = TRUE)
```

#### Part d - For the LASSO models Extract 44th value of lambda and associated coefficients
```{r}
#Extract the 44th value of lambda
lassoModel$lambda[44]

#Calculate the log of the 44th value of lambda (doing this since the plot of coefficient paths used log(lambda))
log(lassoModel$lambda[44])

#Obtain coefficients associated with 44th value of lambda
predict(lassoModel, s = lassoModel$lambda[44], type = "coefficients")

#Add vertical line at 44th lambda on coefficients plot
plot(lassoModel, xvar = "lambda", label = TRUE)
abline(v = log(lassoModel$lambda[44]), col = "red")
```


#### Part e - Pick Optimal Value of Lambda
```{r}
#Use 10-fold CV to pick lambda for LASSO
set.seed(123)
lassoCV <- cv.glmnet(x = XmatTrain, y = yvecTrain,
                    family = "gaussian",
                    alpha = 1, 
                    lambda = NULL,
                    standardize = TRUE,
                    nfolds = 10)
set.seed(NULL)

#Show the LASSO CV results
plot(lassoCV)
```

```{r}
#Use 10-fold CV to pick lambda for Ridge
set.seed(123)
ridgeCV <- cv.glmnet(x = XmatTrain, y = yvecTrain,
                    family = "gaussian",
                    alpha = 0, 
                    lambda = NULL,
                    standardize = TRUE,
                    nfolds = 10)
set.seed(NULL)

#Show the Ridge CV results
plot(ridgeCV)
```

#### Part f - Display the optimal values of lambda and the associated coefficients
```{r}
#Display the optimal values of lambda
lassoCV$lambda.min
lassoCV$lambda.1se

#Store the coefficients associated with the optimal values
coefLamMin <- predict(lassoCV, s = lassoCV$lambda.min, type = "coefficients")
coefLam1se <- predict(lassoCV, s = lassoCV$lambda.1se, type = "coefficients")

#Create a data frame for comparing the coefficients
tempdf <- 
  data.frame(Variable = row.names(coefLamMin), 
             lamMin = as.numeric(coefLamMin), 
             lam1se = as.numeric(coefLam1se))

tempdf
```


#### Part g - Obtain Preditions for Validation Set and calculate RMSE
```{r, eval = FALSE}
#LASSO
lassoYhat <- predict(lassoCV, s = lassoCV$lambda.min, newx = XmatVal) 
lassoMSE <-  mean((yvecVal - lassoYhat)^2)
lassoRMSE <- sqrt(lassoMSE)
lassoRMSE

#Ridge
ridgeYhat <- predict(ridgeCV, s = ridgeCV$lambda.min, newx = XmatVal) 
ridgeMSE <-  mean((yvecVal - ridgeYhat)^2)
ridgeRMSE <- sqrt(ridgeMSE)
ridgeRMSE


#Least Squares (train and validation are not the same as we used in cv.glmnet)
train <- Credit[trainInd, ]
validation <- Credit[-trainInd,]
lsModel <- lm(Balance ~ ., data = train)
lsYhat <- predict(lsModel, newdata = validation)

lsMSE <- mean((validation$Balance - lsYhat)^2)
lsRMSE <- sqrt(lsMSE)
lsRMSE
```

```{r}
library(ISLR2)
library(leaps)
data("Auto")
Auto2 <- Auto[, !(names(Auto) %in% c("origin", "name"))]
Auto2 <- subset(Auto, select = -c(origin, name))
regfit.full <- regsubsets(acceleration ~ ., data = Auto2, nvmax = 7)
reg.summary <- summary(regfit.full)
reg.summary
```

```{r}
library(ISLR2)
library(leaps)

data("Auto")
Auto2 <- subset(Auto, select = -c(origin, name))
regfit.full <- regsubsets(acceleration ~ ., data = Auto2, nvmax = 7)
reg.summary <- summary(regfit.full)
which.min(reg.summary$bic)
reg.summary$which[which.min(reg.summary$bic), ]
```

```





