---
title: "STAT501_L78_Drill_Analysis"
author: "Ada Lazuli"
date: "2025-10-26"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/home/maple/CodeProjects/Statistics/STAT501")
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(qqplotr)
library(tidyr)
library(reactable)
library(openxlsx)
```

# Drill Data Analysis

## Data and Background

A laboratory that tests consumer products selected a variety of drills to compare how likely each will overheat under certain conditions.

The response variable for analysis is **AVTEM** = mean temperature of drill. The dataset includes predictor variables measuring factors to control the drill’s resistance to overheating. 

In particular, **IT** is thickness of insulation, **QW** is an assessment of wire quality in the drill motor, **VS** is the size of the vent used in the drill motor, **I2** is **IT** centered and squared, and **Q2** is **QW** centered and squared.

```{r}
df <- read.csv("~/Downloads/HW6Drill.csv")
head(df)
```

## Modeling

```{r}
model <- lm(AVTEM ~ IT + I2 + VS + QW + Q2, data = df)
summary(model)
```

```{r}
anova(model)
```


Based on the ANOVA output, the sum of squared errors is 1324.2 with degrees of freedom 84

## Analysis of QW and Q2

We need to perform a general linear F-test to determine whether QW and Q2 provide significant information about the response.

$H_0: QW=Q2= 0 \\ H_a: (QW \neq 0) \vee (Q2 \neq 0)$

$$
F^* = \frac{\frac{SSE_R - SSE_F}{P_F - P_R}}{\frac{SSE_F}{n - P_F}}
$$
$$
SSE_R = 1324.2 + 0.3 + 18.2 = 1342.7, \quad P_F = 6,\quad P_R = 4, \quad SSE_F = 1324.2
$$
So,

$$
F^* = \frac{(1342.7-1324.2)/(6-4)}{1324.2/84} = 0.5868
$$

Then, to get the **p-value**,

```{r}
1 - pf(0.5868, 2, 84)
```

Since $P(F_{2,84} \leq x ) = 0.5583671$, we fail to reject the null hypothesis.

At a significance level of 5%, there is not evidence that QW and Q2 contribute significantly after considering the rest of the predictors.

